Data Prep

### 2024-04-03 docs/manuscript.qmd

For this study, we began with data between September 1, 2020 and September 15, 2023 from the \[ornitela\] and \[inpa\] datasets, which we downloaded from Movebank [@kays2022]. For each vulture, we removed any periods of time spent in the wildlife hospital (e.g. while recovering from an illness, injury, or poisoning), or in captivity prior to wild release (in the case of captive-born birds). We cleaned the data by filtering out GPS fixes deemed unreliable (e.g. too few satellites for triangulation, nonsensical heading readings, impossible jumps from previous fixes), and we removed any time that vultures spent in capture cages during the tagging season in the fall. Altogether, approximately `r pctremovedcleaned`% of fixes were removed as compared to the raw data. Next, we removed any vultures that spent fewer than 30 days, in total, within a focal region encompassing the general vicinity of Israel (in order to exclude, for instance, birds that were briefly detected in the area before dispersing far away). For vultures detected in the area for at least 30 days during the study period, we subsequently removed any fixes for those individuals that occurred outside of the same focal region, in order to restrict our analyses to movements that occurred near the main population locus and ensure that most social interactions would be with other tagged individuals. (We accounted for missing time periods, including removed data, by calculating association indices in the social networks).

@@

For each focal individual, we characterized a pattern-based spatial phenotype (Webber et al. 2023) by calculating a variety of movement and space use measures over the course of each of the eight seasons. We used several measures to capture an individual's space use, roosting behavior, and flight movements.

### Space use

We calculated seasonal home ranges and core areas for each focal vulture, using autocorrelated kernel density estimates (AKDEs) via the {ctmm} R package [@fleming2023]. We defined the home range as the 95% AKDE and the core area as the 50% AKDE. To measure differences in how individuals used their home ranges, we calculated the ratio of the seasonal core area size to the seasonal home range size. We defined a general "space use" variable as the first principal component of a PCA containing the home range size and the core area/home range ratio. The "space use" principal component explained \[PERCENT\] of the variation in the data, with individuals low on the principal component having \[CHARACTERISTICS\] and individuals high on the principal component having \[CHARACTERISTICS\].

### Roost diversification

@@

### Movement

We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the {elevatr} package [@hollister2022], at a zoom level of \[level--more details about resolution?\] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, since some tags were occasionally set to a much higher fix rate (e.g. every 1 minute) for short periods of time, we downsampled the data to a 10 minute fix rate to make sure that all individuals had comparable fix rates.

We identified in-flight fixes (ground speed \>= 5 m/s) and calculated measures of daily flight behavior, including maximum and median daily flight altitude and daily flight time (the sum of time intervals between adjacent in-flight fixes). Finally, we used all fixes to calculate daily distance traveled (the sum of all distances between successive fixes); daily maximum displacement (the straight-line distance between the beginning of the daily movement track and the farthest point reached on that day); and movement tortuosity (cumulative distance to the point of maximum displacement divided by the straight-line distance to the point of maximum displacement [@benhamou2004]). We took seasonal means of each of these individual daily measures to get a single value per individual vulture, per season.

## Summarizing movement measures

To summarize individuals' spatial phenotype, we created three composite indices: **movement**, **space use**, and **roost diversification**. Each index was derived from a principal component analysis of several spatial phenotype variables, all of which were scaled and centered over the entire date range of the study before putting them into the PCA.

**Movement** was defined as the first principal component of a PCA containing per-vulture seasonal means of daily mean flight altitude, daily flight duration, daily maximum displacement, and daily flight tortuosity. **Space use** was defined as the first principal component of a PCA containing log-transformed 95% AKDE home range sizes and the ratio of the 50% AKDE to the 95% AKDE. **Roost diversification** was defined as the first principal component of a PCA containing a seasonal Shannon diversity index of roost sites used \[ref Shannon\], number of unique roost sites used, and the proportion of nights in a given season when a vulture switched to a new roost site (versus roosting at the same site as on the previous night). The composite measures accounted for %, %, and % of the variation in their respective PCAs.

@@

Using GPS fixes from the tagged vultures, we created proximity-based social networks in three different social situations, following Sharma [-@sharma2023]: co-flight, co-feeding (diurnal ground interactions) and co-roosting (nocturnal ground interactions). Networks were created separately for each of the eight seasons. For each social situation in each season, we created a weighted, undirected network. Because not all individuals were tagged for exactly the same period of time, and to account for missing data due to low tag battery or poor reception, we calculated the simple ratio index (SRI) to quantify the strength of the association between each pair of vultures in each social situation. We calculated SRI by dividing the number of 10-minute time periods during which two vultures were observed interacting by the number of time periods during which both vultures had a recorded GPS fix (whether or not they were interacting during that period) [@ginsberg1992]. Thus, two vultures who never interacted in a given social situation would have an SRI of 0, while two individuals who were interacting in every time period during which a GPS fix was collected for both of them would have an SRI of 1.

@@

Co-feeding

Vultures that were not flying (ground speed \< 5 m/s, [@spiegel2013]) who came within 50m of each other for at least two consecutive 10-minute time periods, were considered to be having a ground interaction [@sharma2023]. Any ground interactions that occurred at night or within the boundaries of a known roost site were excluded. Therefore, the remaining diurnal ground interactions should represent vultures' interactions while feeding or while present at feeding sites (Orr Spiegel and Nili Anglister, pers. comm.). Co-roosting Roost sites were assigned using [methods]. Unlike flight and diurnal ground GPS fixes, vultures were only assigned one roosting location per night. Two vultures that roosted within the boundaries of the same roost site polygon on a given night were considered to be co-roosting.

@@

We modeled the relationship between the spatial phenotype variables and the network measures using linear mixed-effects regression models, fitted using the {lme4} and {lmerTest} packages [@bates2015; @kuznetsova2017]. We fit separate models for each response variable (degree and strength). The predictors included were movement (PC1), exploration (PC2), situation (feeding, flight, roosting), age group (juvenile/subadult, adult), and season (breeding, summer, fall). To account for the different numbers of individuals in the network in each unique season, we included a random effect for the interaction of season and year, e.g. "2022_fall" (8 levels). To account for consistent individual differences in network position or behavior, and because individuals varied in how many seasons they were included in, we also included an individual-level random effect.

@@

```{r}
#| eval: false
source(here("scripts/04.0_mixedModels.R"))
```

```{r}
#| eval: false
#| label: fig-d-space
#| fig-cap: "Relationship between normalized degree and space use"
plot_d_eff_space
```

```{r}
#| eval: false
#| label: fig-d-roost
#| fig-cap: "Relationship between normalized degree and roost diversification"
plot_d_eff_roost
```

```{r}
#| eval: false
#| label: fig-d-space-situ
#| fig-cap: "Relationship between normalized degree and space use, with separate trend lines for each social situation"
plot_d_eff_situ_space
plot_d_emt_situ_space
```

```{r}
#| eval: false
#| label: fig-s-movement
#| fig-cap: "Relationship between normalized strength and movement"
plot_s_eff_movement
```

```{r}
#| eval: false
#| label: fig-d-movement-situ
#| fig-cap: "Relationship between normalized degree and movement, with separate trend lines for each social situation"
plot_d_eff_situ_movement
plot_d_emt_situ_movement
```

@@

```{r}
#| eval: false
#| label: fig-s-space
#| fig-cap: "Relationship between normalized strength and space use"
plot_s_eff_situ_space
plot_s_emt_situ_space
```

>>>>

For this study, we used data between September 1, 2020 and September 15, 2023 from the \[ornitela\] and \[inpa\] datasets, which we downloaded from Movebank [@kays2022]. For each vulture, we removed any periods of time spent in the wildlife hospital (e.g. while recovering from an illness, injury, or poisoning), or in captivity prior to wild release (in the case of captive-born birds). We cleaned the data by filtering out GPS fixes deemed unreliable (e.g. too few satellites for triangulation, nonsensical heading readings, impossible jumps from previous fixes), and we removed any time that vultures spent in capture cages during the tagging season in the fall. Altogether, approximately `r pctremovedcleaned`% of fixes were removed as compared to the raw data. Next, we removed any vultures that spent fewer than 30 days, in total, within a focal region encompassing the general vicinity of Israel (in order to exclude, for instance, birds that were briefly detected in the area before dispersing far away). For vultures detected in the area for at least 30 days during the study period, we subsequently removed any fixes for those individuals that occurred outside of the same focal region, in order to restrict our analyses to movements that occurred near the main population locus and ensure that most social interactions would be with other tagged individuals. (We accounted for missing time periods, including removed data, by calculating association indices in the social networks).

@@

We characterized the pattern-based spatial phenotype [@webber2023a] of each focal individual by measuring its overall space use, use of roost sites, and movement patterns.

We calculated seasonal home ranges and core areas using autocorrelated kernel density estimates (AKDEs) via the {ctmm} R package [@fleming2023]. We defined the home range as the 95% AKDE and the core area as the 50% AKDE. To measure differences in how individuals used their home ranges, we calculated the ratio of the seasonal core area size to the seasonal home range size.

@@

We identified in-flight fixes (ground speed \>= 5 m/s) and calculated measures of daily flight behavior, including maximum and median daily flight altitude and daily flight time (the sum of time intervals between adjacent in-flight fixes). We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the {elevatr} package at a ground resolution of approximately 250 meters [@hollister2022] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, we used all fixes to calculate daily distance traveled (the sum of all distances between successive fixes); daily maximum displacement (the straight-line distance between the beginning of the daily movement track and the farthest point reached on that day); and movement tortuosity (cumulative distance to the point of maximum displacement divided by the straight-line distance to the point of maximum displacement [@benhamou2004]). We took seasonal means of each of these individual daily measures to get a single value per individual, per season.

## Summarizing movement measures
```{r}
#| include: false
load(here("data/calcMovement/pca_space.Rda"))
propvar_space <- as.data.frame(summary(pca_space)$importance)$PC1[2]

load(here("data/calcMovement/pca_roost.Rda"))
propvar_roost <- as.data.frame(summary(pca_roost)$importance)$PC1[2]

load(here("data/calcMovement/pca_movement.Rda"))
propvar_movement <- as.data.frame(summary(pca_movement)$importance)$PC1[2]
```

We characterized individuals' spatial phenotypes using three composite indices: movement, space use, and roost diversification. Each index was derived from a principal component analysis of several spatial phenotype variables, each scaled and centered over the entire date range of the study.

**Movement** was defined as the first principal component of a PCA containing per-vulture seasonal means of daily mean flight altitude, daily flight duration, daily maximum displacement, and daily flight tortuosity. It explained `r 100*round(propvar_movement,2)`% of the variation in the PCA. Individuals with high movement scores traveled farther and flew longer, with straighter daily trajectories overall. **Space use** was defined as the first principal component of a PCA containing log-transformed 95% AKDE home range sizes and the ratio of the 50% AKDE to the 95% AKDE. It explained `r 100*round(propvar_space,2)`% of the variation in the PCA. Individuals low on space use had small overall home ranges with concentrated core areas (and some exploratory movements away from the core), while individuals high on space use had large home ranges consisting of space that they used frequently (i.e. the core area made up a larger proportion of the home range). **Roost diversification** was defined as the first principal component of a PCA containing a seasonal Shannon diversity index of roost sites used \[ref Shannon\], number of unique roost sites used, and the proportion of nights in a given season when a vulture switched to a new roost site (versus roosting at the same site as on the previous night). It explained `r 100*round(propvar_roost,2)`% of the variation in the PCA.

While the griffon vulture population in southern Israel is fairly well mixed and vultures do not have discrete, exclusive territories, there is still variability in where individuals spend their time within the landscape. Spatial position has the potential to directly affect how many other individuals a vulture encounters and which feeding and roost sites it frequents. In addition, though a high proportion of the vultures in the area are GPS-tagged, more peripheral vultures might be more likely to interact with untagged birds living farther from the population center, potentially diluting their measured measured strength or degree. To account for the influence of geographic centrality on social centrality, we calculated the geographic centroid of all GPS points for each individual in each season. Then we compared these centroids to the overall centroid of all individuals in the population in that season to obtain an index of centrality (versus peripherality). We incorporated spatial centrality as a covariate into our linear models.

@@

Using GPS fixes from the tagged vultures, we created weighted, undirected proximity-based social networks in three different social situations for each of the nine seasons, following Sharma [-@sharma2023]: co-flight, co-feeding (diurnal ground interactions) and co-roosting (nocturnal ground interactions). Because not all individuals were tagged for exactly the same period of time, and to account for missing data due to low tag battery or poor reception, we calculated the simple ratio index (SRI) to quantify the strength of the association between each pair of vultures in each social situation. We calculated SRI by dividing the number of 10-minute time periods during which two vultures were observed interacting by the number of time periods during which both vultures had a recorded GPS fix (whether or not they were interacting during that period) [@ginsberg1992]. Thus, two vultures who never interacted in a given social situation would have an SRI of 0, while two individuals who were interacting in every time period during which a GPS fix was collected for both of them would have an SRI of 1.

@@

### Co-feeding

Vultures that were not flying (ground speed \< 5 m/s, [@spiegel2013]) who came within 50m of each other for at least two consecutive 10-minute time periods, were considered to be having a ground interaction [@sharma2023]. Any ground interactions that occurred at night or within the boundaries of a known roost site were excluded. Therefore, the remaining diurnal ground interactions should represent vultures' interactions while feeding or while present at feeding sites (Orr Spiegel and Nili Anglister, pers. comm.).

### Co-roosting 
Vultures were considered to be co-roosting if they roosted within 500m of each other on a given night. Unlike interactions in the co-feeding and co-flight networks, co-roosting interactions could occur at a maximum frequency of once per night. Therefore, edge weights in the co-roosting network represented the number of nights on which two individuals co-roosted divided by the number of nights when both individuals had a valid roost location (i.e. they theoretically could have co-roosted), regardless of proximity.

@@

We modeled the relationship between the spatial phenotype variables and the network measures using linear mixed-effects regression models, fitted using the {lme4} and {lmerTest} packages [@bates2015; @kuznetsova2017]. We fit separate models for each response variable (degree and strength). The predictors included were movement (PC1), exploration (PC2), situation (feeding, flight, roosting), age group (juvenile/subadult, adult), season (breeding, summer, fall), and geographic centrality. To account for the different numbers of individuals in the network in each unique season, we included a random effect for the interaction of season and year, e.g. "2022_fall" (8 levels). To account for consistent individual differences in network position or behavior, and because individuals varied in how many seasons they were included in, we also included an individual-level random effect.

@@

```{r}
#| include: false
#| output: false
source(here("scripts/04.0_mixedModels.R"))
```

## Degree

```{r}
#| eval: false
#| output: false
#| label: fig-d-space
#| fig-cap: "Relationship between normalized degree and space use"
# NOT SHOW
plot_d_eff_space # null--no overall relationship between space use and degree
```

```{r}
#| label: fig-d-space-situ
#| fig-cap: "Relationship between normalized degree and space use, with separate trend lines for each social situation"
plot_d_eff_situ_space
plot_d_emt_situ_space
```
There was a significant relationship between how vultures used space and how many other individuals they encountered in flight. Vultures that consistently ranged widely encountered more others in flight, while those that concentrated their movement in smaller home ranges and smaller proportional core areas flew with fewer others. There was a slight, albeit non-statistically-significant, trend toward lower degree in the co-feeding and co-roosting networks for vultures with larger ranges.

```{r}
#| eval: false
#| output: false
#| label: fig-d-roost
#| fig-cap: "Relationship between normalized degree and roost diversification"
# NOT SHOW
plot_d_eff_roost # null--no overall relationship between roost diversification and degree
```

```{r}
#| eval: false
#| output: false
#| label: fig-d-movement-situ
#| fig-cap: "Relationship between normalized degree and movement, with separate trend lines for each social situation"
# NOT SHOW
plot_d_eff_situ_movement
plot_d_emt_situ_movement # going to show the plot by season instead of this one
```

```{r}
#| label: fig-d-movement-situ-season
#| fig-cap: "Relationship between normalized degree and movement, for each season and social situation."
plot_d_emt_situ_movement_season
```

Movement behavior affected degree in all three social situations. Vultures with a higher movement score co-roosted with fewer conspecifics only in the transient, pre-breeding season; otherwise movement did not affect their co-roosting degree. In the breeding and post-breeding seasons, vultures that moved more encountered more conspecifics in flight and while feeding, but this relationship did not hold in the pre-breeding season.

## Strength

```{r}
#| eval: false
#| output: false
#| label: fig-s-movement
#| fig-cap: "Relationship between normalized strength and movement"
plot_s_eff_movement # null
```

There was no relationship between movement and strength, either overall or in any season or situation.

@@

However, individuals with more diversified roosts had lower strength (i.e. they interacted less) overall, considering seasons and situations together.

```{r}
#| eval: false
#| label: fig-s-space
#| fig-cap: "Relationship between normalized strength and space use"
plot_s_eff_situ_space
plot_s_emt_situ_space
```
Vultures with greater space use had significantly more co-flight interactions (higher strength in the co-flight network). They also tended to have a higher co-roosting strength, though the trend was not statistically significant. There was no relationship between seasonal space use and co-feeding strength.

### 2024-04-03 scripts/00.0_dataPrep.qmd

## Downsample to 5-minute frequency
This data is just WAY too huge to handle now that we have the higher-frequency fixes from the summer/fall of 2023. Going to downsample it a bit.

@@

```{r remove-periods}
removed_periods <- vultureUtils::removeInvalidPeriods(dataset = interval_5min, periodsToRemove = periods_to_remove)
write_feather(removed_periods, here("data/dataPrep/removed_periods.feather"))
```

@@

```{r}
# warning: this takes a long time and uses a lot of memory!
data_cleaned <- vultureUtils::cleanData(dataset = removed_periods,
                     precise = F,
                     longCol = "location_long",
                     latCol = "location_lat",
                     idCol = "Nili_id",
                     report = F)
rm(joined)
gc()
save(data_cleaned, file = here("data/dataPrep/data_cleaned.Rda"))

(nrow(removed_periods)-nrow(data_cleaned))/nrow(removed_periods) # removed less than 1% of the points.
```

## Remove time spent in capture cages
```{r}
capture_sites <- read.csv(here("data/raw/capture_sites.csv"))
carmel <- read.csv(here("data/raw/all_captures_carmel_2010-2021.csv"))
removed_captures <- removeCaptures(data = data_cleaned, 
                                       captureSites = capture_sites, 
                                       AllCarmelDates = carmel, 
                                       distance = 500, idCol = "Nili_id")

(nrow(removed_captures)-nrow(data_cleaned))/nrow(data_cleaned) # did not remove any data. Should probably check on this...
save(removed_captures, file = here("data/dataPrep/removed_captures.Rda"))
```

@@

@@

# Region masking
data_masked <- vultureUtils::inMaskFilter(dataset = with_age_sex, mask = mask, inMaskThreshold = 0.33, crs = "WGS84") # this takes a long time!!
save(data_masked, file = here("data/dataPrep/data_masked.Rda")) # note that this saves a 3-element list. Need to get Ryan to fix this.#XXX
load(here("data/dataPrep/data_masked.Rda"))
length(data_masked$Nili_id)
data_masked_2 <- data_masked$dataset

# Fix time zone so dates make sense ---------------------------------------
## Overwrite the dateOnly column from the new times
data_masked_2 <- data_masked_2 %>%
  mutate(timestampIsrael = lubridate::with_tz(timestamp, tzone = "Israel"),
         dateOnly = lubridate::date(timestampIsrael),
         month = lubridate::month(dateOnly),
         year = lubridate::year(dateOnly))
save(data_masked_2, file = here("data/dataPrep/data_masked_2.Rda")) # just the df #XXX fix

@@

```{r split-seasons}
load(here("data/dataPrep/data_masked_2.Rda")) # XXX fix

# Split into Marta's 3 seasons --------------------------------------------
split <- data_masked_2 %>%
  mutate(start_breeding = ifelse(month %in% c(1:5),
                                 paste(as.character(year-1), "-", "12", "-", "15", sep = ""),
                                 paste(as.character(year), "-", "12", "-", "15", sep = "")),
         start_summer = ifelse(month == 12,
                               paste(as.character(year + 1), "-", "05", "-", "15", sep = ""),
                               paste(as.character(year), "-", "05", "-", "15", sep = "")),
         start_fall = as.Date(paste(as.character(year), "-", "09", "-", "15", sep = "")),
         start_breeding = as.Date(start_breeding),
         start_summer = as.Date(start_summer),
         season = case_when(
           dateOnly >= start_breeding & dateOnly < start_summer ~ "breeding",
           dateOnly >= start_summer & dateOnly <= start_fall ~ "summer",
           dateOnly >= start_fall & dateOnly < start_breeding ~ "fall")) %>%
  filter(dateOnly != "2023-09-15") # remove September 15 2023, because it turns out the season boundaries are supposed to be non-inclusive and this is the latest one.

with_seasons <- split %>%
  mutate(age = year - birth_year, #XXX consider moving this up to the age/sex section. Why is it happening here?
         sex = case_when(sex == "f/m" ~ "u",
                         TRUE ~ sex)) %>%
  mutate(age = ifelse(season == "breeding" & month %in% c(1, 12), age + 1, age)) %>%
  group_by(Nili_id, year, month) %>%
  mutate(age = ifelse(month == 2, min(age) + 1, age)) %>%
  mutate(age_group = case_when(age == 0 ~ "juv",
                               age >= 1 & age <= 4 ~ "sub",
                               age >= 5 ~ "adult",
                               TRUE ~ NA)) %>%
  mutate(seasonUnique = case_when(season %in% c("fall", "summer") ~ paste(year, season, sep = "_"),
                                  season == "breeding" & month == 12 ~ paste(year + 1, season, sep = "_"),
                                  season == "breeding" & month != 12 ~ paste(year, season, sep = "_"))) %>%
  # explicitly set the factor levels in temporal order
  mutate(seasonUnique = factor(seasonUnique, levels = c("2020_summer", "2020_fall", "2021_breeding", "2021_summer", "2021_fall", "2022_breeding", "2022_summer", "2022_fall", "2023_breeding", "2023_summer")),
         season = factor(season, levels = c("breeding", "summer", "fall")))

@@

```{r}
load(here::here("data", "dataPrep", "removed_northern.Rda"))
load(here::here("data", "dataPrep", "season_names.Rda"))
```

@@

```{r}
removed_lfr <- map2(removed_northern, low_fix_rate_indivs,                     ~.x %>% filter(!(Nili_id %in% .y))) 
save(removed_lfr, file = here::here("data", "dataPrep", "removed_lfr.Rda"))  
# tidy up 
rm(removed_northern) 
rm(low_fix_rate_indivs) 
rm(removed_lfr) 
gc()
```

@@

```{r}
#| eval: false
downsampled_10min_forSocial <- map(removed_lfr, ~subsample(df = .x, idCol = "Nili_id", timestampCol = "timestamp", mins = 10), .progress = T)

# Add information on # of days tracked for each individual
downsampled_10min_forSocial <- map(downsampled_10min_forSocial, ~.x %>% group_by(Nili_id) %>% 
                       mutate(daysTracked = length(unique(dateOnly))))
save(downsampled_10min_forSocial, file = here::here("data", "dataPrep", "downsampled_10min_forSocial.Rda"))

map_dbl(removed_lfr, nrow) %>% setNames(season_names)
map_dbl(downsampled_10min_forSocial, nrow) %>% setNames(season_names)

# tidying
rm(removed_lfr)
rm(downsampled_10min_forSocial)
gc()
```

@@

```{r}
load(here::here("data", "dataPrep", "downsampled_10min_forSocial.Rda"))
```

@@

# Tidy up
rm(downsampled_10min_forSocial)
rm(roosts)
gc()
```

@@

```{r}
load(here::here("data", "dataPrep", "removed_lfr.Rda"))
load(here::here("data", "dataPrep", "season_names.Rda"))
```

@@

save(removed_nighttime, file = here::here("data", "dataPrep", "removed_nighttime.Rda"))

# tidy up
rm(removed_nighttime)
rm(removed_lfr)
rm(season_names)
gc()
```

@@

```{r}
load(here::here("data", "dataPrep", "removed_nighttime.Rda"))
```

```{r}
#| eval: false
# Calculate number of points per day for each individual
ppd <- map_dfr(removed_nighttime, ~.x %>%
                 st_drop_geometry() %>%
                 group_by(seasonUnique, season, Nili_id, dateOnly) %>%
                 summarize(n = n(),
                           minBatt = min(battery_charge_percent, na.rm = T)), .progress = T)

@@

# tidying up
rm(removed_northern)
rm(ppd)
rm(excl_batt)
rm(excl_pts)
rm(removed_lowppd)
gc()
```

@@

# defined the "subsample" function previously above

# create and save downsampled datasets for later use
downsampled_10min <- map(with_altitudes, ~subsample(df = .x, idCol = "Nili_id", timestampCol = "timestamp", mins = 10), .progress = T) # basically just trying to remove any data from the occasional bursts of more concentrated and frequent sampling.

# Add information on # of days tracked for each individual
downsampled_10min <- map(downsampled_10min, ~.x %>% group_by(Nili_id) %>% 
                       mutate(daysTracked = length(unique(dateOnly))))

save(downsampled_10min, file = here::here("data", "dataPrep", "downsampled_10min.Rda"))

map_dbl(with_altitudes, nrow) %>% setNames(season_names)
map_dbl(downsampled_10min, nrow) %>% setNames(season_names)

>>>>

## Downsample to 5-minute frequency

This data is just WAY too huge to handle now that we have the higher-frequency fixes from the summer/fall of 2023. Going to downsample it a bit.

@@

```{r remove-periods}
removed_periods <- vultureUtils::removeInvalidPeriods(dataset = interval_5min, periodsToRemove = periods_to_remove)
save(removed_periods, file = here("data/dataPrep/removed_periods.Rda"))
```

@@

```{r}
# warning: this takes a long time and uses a lot of memory!
cleaned <- vultureUtils::cleanData(dataset = removed_periods,
                     precise = F,
                     longCol = "location_long",
                     latCol = "location_lat",
                     idCol = "Nili_id",
                     report = F)
rm(interval_5min)
gc()
save(cleaned, file = here("data/dataPrep/cleaned.Rda"))

(nrow(removed_periods)-nrow(cleaned))/nrow(removed_periods) # removed less than 1% of the points.
```

## Remove time spent in capture cages

```{r}
capture_sites <- read.csv(here("data/raw/capture_sites.csv"))
carmel <- read.csv(here("data/raw/all_captures_carmel_2010-2021.csv"))
removed_captures <- removeCaptures(data = cleaned, 
                                       captureSites = capture_sites, 
                                       AllCarmelDates = carmel, 
                                       distance = 500, idCol = "Nili_id")

(nrow(removed_captures)-nrow(cleaned))/nrow(cleaned) # did not remove any data. Should probably check on this...
save(removed_captures, file = here("data/dataPrep/removed_captures.Rda"))
```

@@

mutate(age = year - birth_year, #XXX consider moving this up to the age/sex section. Why is it happening here?
         sex = case_when(sex == "f/m" ~ "u",
                         TRUE ~ sex)) %>%

@@

# Region masking
data_masked <- vultureUtils::inMaskFilter(dataset = with_age_sex, mask = mask, inMaskThreshold = 0.33, crs = "WGS84")$dataset # this takes a long time!!

# Fix time zone so dates make sense ---------------------------------------
## Overwrite the dateOnly column from the new times
data_masked <- data_masked %>%
  mutate(timestampIsrael = lubridate::with_tz(timestamp, tzone = "Israel"),
         dateOnly = lubridate::date(timestampIsrael),
         month = lubridate::month(dateOnly),
         year = lubridate::year(dateOnly))
save(data_masked, file = here("data/dataPrep/data_masked.Rda")) # just the df #XXX fix

@@

```{r split-seasons}
load(here("data/dataPrep/data_masked.Rda")) # XXX fix
data_masked <- sf::st_drop_geometry(data_masked) # to make it faster to work on

# Split into Marta's 3 seasons --------------------------------------------
with_seasons <- data_masked %>% mutate(month = lubridate::month(timestampIsrael),
                day = lubridate::day(timestampIsrael),
                year = lubridate::year(timestampIsrael)) %>%
  mutate(season = case_when(((month == 12 & day >= 15) | 
                              (month %in% 1:4) | 
                              (month == 5 & day < 15)) ~ "breeding",
                            ((month == 5 & day >= 15) | 
                               (month %in% 6:8) | 
                               (month == 9 & day < 15)) ~ "summer",
                            .default = "fall")) %>%
  filter(dateOnly != "2023-09-15") %>% # remove September 15 2023, because it turns out the season boundaries are supposed to be non-inclusive and this is the latest one.
  mutate(seasonUnique = case_when(season == "breeding" & month == 12 ~
                                    paste(as.character(year + 1), season, sep = "_"),
                                  .default = paste(as.character(year), season, sep = "_"))) %>%
  mutate(seasonUnique = factor(seasonUnique, levels = c("2020_summer", "2020_fall", "2021_breeding", "2021_summer", "2021_fall", "2022_breeding", "2022_summer", "2022_fall", "2023_breeding", "2023_summer")),
         season = factor(season, levels = c("breeding", "summer", "fall")))

# Add ages (based on season, which is why this goes here rather than above)
with_seasons <- with_seasons %>%
  mutate(age = year - birth_year) %>%
  mutate(age = ifelse(season == "breeding" & month %in% c(1, 12), age + 1, age)) %>%
  group_by(Nili_id, year, month) %>%
  mutate(age = ifelse(month == 2, min(age) + 1, age)) %>%
  mutate(age_group = case_when(age == 0 ~ "juv",
                               age >= 1 & age <= 4 ~ "sub",
                               age >= 5 ~ "adult",
                               .default = NA)) %>%
  ungroup()

@@

```{r}
load(here::here("data/dataPrep/removed_northern.Rda"))
load(here::here("data/dataPrep/season_names.Rda"))
```

@@

```{r}
removed_lfr <- map2(removed_northern, low_fix_rate_indivs,                     
                    ~.x %>% filter(!(Nili_id %in% .y))) 
save(removed_lfr, file = here::here("data", "dataPrep", "removed_lfr.Rda"))  
# tidy up 
rm(removed_northern) 
rm(low_fix_rate_indivs) 
gc()
```

## Remove high-frequency fixes (downsample for social)

@@

```{r}
lst <- map(removed_lfr, ~.x %>% group_split(Nili_id))
downsampled_10min_forSocial <- map(lst, ~{
  purrr::list_rbind(map(.x, ~down(df = .x, mins = 10), .progress = T))
}, .progress = T)

# Add information on # of days tracked for each individual
downsampled_10min_forSocial <- map(downsampled_10min_forSocial, ~.x %>% group_by(Nili_id) %>% 
                       mutate(daysTracked = length(unique(dateOnly))))
save(downsampled_10min_forSocial, file = here::here("data/dataPrep/downsampled_10min_forSocial.Rda"))

# tidying
rm(removed_lfr)
gc()
```

@@

```{r}
load(here::here("data/dataPrep/downsampled_10min_forSocial.Rda"))
```

@@

# Tidy up
rm(downsampled_10min_forSocial)
gc()
```

@@

```{r}
load(here::here("data/dataPrep/removed_lfr.Rda"))
load(here::here("data/dataPrep/season_names.Rda"))
```

@@

save(removed_nighttime, file = here::here("data/dataPrep/removed_nighttime.Rda"))

# tidy up
rm(removed_lfr)
rm(season_names)
gc()
```

@@

```{r}
load(here::here("data/dataPrep/removed_nighttime.Rda"))
```

```{r}
#| eval: false
# Calculate number of points per day for each individual
# XXX START HERE can't find battery charge percent
ppd <- map_dfr(removed_nighttime, ~.x %>%
                 st_drop_geometry() %>%
                 group_by(seasonUnique, season, Nili_id, dateOnly) %>%
                 summarize(n = n(),
                           minBatt = min(battery_charge_percent, na.rm = T)) %>% 
                 ungroup(), .progress = T)

@@

# tidying up
rm(removed_nighttime)
rm(removed_northern)
rm(ppd)
rm(excl_batt)
rm(excl_pts)
gc()
```

@@

```{r}
lst_2 <- map(with_altitudes, ~.x %>% group_split(Nili_id))
downsampled_10min <- map(lst_2, ~{
  purrr::list_rbind(map(.x, ~down(df = .x, mins = 10), .progress = T))
}, .progress = T)

# Add information on # of days tracked for each individual
downsampled_10min <- map(downsampled_10min, ~.x %>% group_by(Nili_id) %>% 
                       mutate(daysTracked = length(unique(dateOnly))))
save(downsampled_10min, file = here::here("data/dataPrep/downsampled_10min.Rda"))
